---
title: "CFSS Homework 02"
author: "Cory J. Cascalheira"
date: "November 28, 2018"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/GitHub/cfss-hw02")
```

# Exploring Gun Deaths in America
___
Let's explore gun deaths in America between the years 2012 to 2014. These data, originally compiled by FiveThirtyEight, are taken from [MACS 30500](https://cfss.uchicago.edu/index.html), Computing for the Social Sciences, a course taught by [Ben Soltoff](http://www.bensoltoff.com/) at the University of Chicago.

TO begin, we must load dependencies and import data to construct the workspace.

```{r message=FALSE}
library(tidyverse) # utility & visualization
library(knitr) # tables
library(kableExtra) # awesome tables
library(RColorBrewer) # custome colors

# Import data
load(file = "data/gun_deaths.rda")
```

Let's also set the default theme for all of our plots.

```{r}
theme_set(theme_classic())
```

## Understand the Variables
___
If the data were untidy, now would be the time to clean them. That is, we would want to ensure that each cell has exactly one value, each column one variable, and each row one observation. Thanfully, the instructor has cleaned the data ahead of time.

```{r}
glimpse(gun_deaths)
```

Aside from one `double` and `factor`, the data are comprised of `integer`s and `character`s.

What are the levels of the factor?

```{r}
levels(gun_deaths$education)
```

Alright, so the `education` variable is at the ordinal level.

## Number of Gun Deaths per Month
___
How many gun deaths occurred each month in the United States over the course of two years?

Before we tabulate and visualize the data, let's transform `month` into a factor with named levels.

```{r}
# Vector of month names
abbr_months <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")

# Transform month into a named factor
gun_deaths2 <- within(gun_deaths, {
  month <- factor(month, labels = abbr_months) 
})
```

Tables exhibit quantitative information more precisely than graphics. Algorithms prefer tabular data.

```{r}
# Count the rows, which represent deaths
count(gun_deaths2, month) %>%
  # Create the HTML table
  kable(caption = "Total Gun Deaths in the United States (2012-2014) per Month",
        col.names = c("Month", "Total Deaths")) %>%
  # Give the table some CSS love
  kable_styling(bootstrap_options = c("striped", "hover"))
```

From the table[^1], we see that the number of deaths appears constant. Aside from February, the total number of deaths for each month is above 8,000. It is important to remember that the table aggregates deaths for both years by month.

As humans, we tend to grasp patterns in visualizations better. The only number that jumps out to us after a quick glance at the above table is the leading digital, eight. Our minds blur the remaining digits as nonsensical. This is why $5.99 feels like five dollars and not six.

What does a bar chart[^2] tell us?

```{r}
ggplot(gun_deaths2, aes(x = month)) +
  geom_bar() +
  # Describe the graph
  labs(title = "Gun Deaths in the United States",
       subtitle = "From 2012 to 2014",
       caption = "Source: FiveThirtyEight",
       x = "Month",
       y = "Number of Deaths") +
  # Force origin to start at 0
  scale_y_continuous(expand = c(0, 0))
```

It is tempting to think that the fewest deaths occurred in February due to Valentine's Day. Romantic! However, perhaps this is due to the shortness of the month--fewer days to commit homicide.

This pattern holds even when we break down the months by year.

```{r}
ggplot(gun_deaths2, aes(x = month)) +
  geom_bar() +
  facet_wrap(~ year) +
  # Rotate axis labels
  theme(axis.text.x = element_text(angle = 45))
```

## Number of Gun Deaths by Intent
___
Earlier, I hypothesized that there are fewer deaths in the month of February due to fewer days to commit homicide. But maybe people kill themselves in February due to loneliness?

```{r}
unique(gun_deaths2$intent)
```

Intent is an important variable to consider. First, how many gun deaths are associated with specific intents?

```{r}
ggplot(gun_deaths2, aes(x = intent)) +
  geom_bar()
```

Unfortunately, suicide is the leading cause of gun death in the United States, at least between the years 2012 and 2014.

Now that we know how many deaths correspond to each type of intent, we can reorder the bar graph. Let's remove the `NA` values, too. Reordering a bar graph requires a transformation of `intent` into a factor and specification of the levels.

```{r}
# Intent as a factor
gun_deaths3 <- within(gun_deaths2, {
  intent <- factor(intent, levels = c("Suicide", "Homicide", "Accidental", "Undetermined", NA))
})

# Remove NA
gun_deaths3 %>%
  filter(!is.na(intent)) %>%
  # Plot reordered bar graph
  ggplot(aes(x = intent)) +
    geom_bar() +
    labs(title = "Cause of Gun Deaths",
         x = "Intent",
         y = "Number of Deaths")
```

So, are there more suicides in February?

```{r}
gun_deaths3 %>%
  filter(!is.na(intent)) %>%
  ggplot(aes(x = month, fill = intent)) +
    geom_bar(position = "fill") +
    # Draw horizontal line to compare suicide across months
    geom_hline(yintercept = 0.333) +
    scale_fill_brewer(palette = "Spectral")
```

Yes, they are.

## Age of Victims by Sex
___
How old are the victims when they die by gun violence?

```{r warning=FALSE}
ggplot(gun_deaths3, aes(x = sex, y = age)) +
  geom_boxplot()
```

```{r}
(gun_death_desc <- gun_deaths3 %>% 
  group_by(sex) %>%
  summarize(
    percentile_75 = quantile(age, p = .75,  na.rm = TRUE),
    median_age = median(age, na.rm = TRUE),
    percentile_25 = quantile(age, p = 0.25, na.rm = TRUE),
    mean_age = mean(age, na.rm = TRUE)
  ))
```

Most victims die in the middle of life, men between the ages of `r gun_death_desc[2, 4]` and `r gun_death_desc[2, 2]`. Women die most often around the age of `r round(gun_death_desc[1, 5])`.

## Gun Death of White Males in 2012
___
How many white males with at least a high school education were killed by guns in 2012?

We can answer this question by filtering with logical statements. Note that at least a high school education is the same as not less than a high school education.

```{r}
levels(gun_deaths3$education)
```

```{r}
gun_deaths3 %>%
  # Filter rows for target demographic in appropriate year
  filter(education != "Less than HS", race == "White", sex == "M", year == 2012) %>%
  count()
```

15,199 White men with at least a high school education died by gun violence in 2012. 

Delineation of this quantity may reveal other trends.

```{r}
gun_deaths3 %>%
  # Filter rows for target demographic in appropriate year
  filter(education != "Less than HS", race == "White", sex == "M", year == 2012) %>%
  ggplot(aes(x = intent)) +
    geom_bar(position = "dodge") +
    scale_y_continuous(expand = c(0, 0)) +
    labs(title = "Cause of Death by Gun Violence for Educated White Males (2012)",
         x = "Cause of Death",
         y = "Number of Deaths")
```

Suicide is the leading cause of gun-related death for this demographic.

## Gun Deaths by Season
___
Which season of the year has the most gun deaths?

We can use `cut()` on the original data frame to divide `month` into four equal seasons.

```{r}
ggplot(gun_deaths, aes(x = cut(month, 4, include.lowest = TRUE,
                               labels = c("Winter", "Spring", "Summer", "Fall")))) +
  geom_bar() +
  labs(x = "Season",
       y = "Number of Deaths")
```

Summer has the highest number of gun-related deaths.

## Death Disparities by Race
___
Are whites who are killed by guns more likely to die because of suicide or homicide? How does this compare to blacks and hispanics?

## Police-Involved Gun Deaths
___
Are police-involved gun deaths significantly different from other gun deaths?

## Dependencies
___
This activity was completed in RStudio.

```{r}
devtools::session_info()
```

[^1]: For more information on how to create awesome tables, check out Hao Zhu's [documentation](https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html).
[^2]: Start the [origin at zero](https://stackoverflow.com/questions/13701347/force-the-origin-to-start-at-0).