---
title: "CFSS Homework 02"
author: "Cory J. Cascalheira"
date: "November 28, 2018"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/GitHub/cfss-hw02")
```

# Exploring Gun Deaths in America
___
Let's explore gun deaths in America between the years 2012 to 2014. These data are taken from [MACS 30500](https://cfss.uchicago.edu/index.html), Computing for the Social Sciences, a course taught by [Ben Soltoff](http://www.bensoltoff.com/) at the University of Chicago.

TO begin, we must load dependencies and import data to construct the workspace.

```{r message=FALSE}
library(tidyverse) # utility & visualization
library(knitr) # tables

# Import data
load(file = "data/gun_deaths.rda")
```

## Understand the Variables
___
If the data were untidy, now would be the time to clean them. That is, we would want to ensure that each cell has exactly one value, each column one variable, and each row one observation. Thanfully, the instructor has cleaned the data ahead of time.

```{r}
glimpse(gun_deaths)
```

Aside from one `double` and `factor`, the data are comprised of `integer`s and `character`s.

What are the levels of the factor?

```{r}
levels(gun_deaths$education)
```

Alright, so the `education` variable is at the ordinal level.

## Dependencies
___
This activity was completed in RStudio.

```{r}
devtools::session_info()
```

