---
title: "CFSS Homework 02"
author: "Cory J. Cascalheira"
date: "November 28, 2018"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/GitHub/cfss-hw02")
```

# Exploring Gun Deaths in America
___
Let's explore gun deaths in America between the years 2012 to 2014. These data are taken from [MACS 30500](https://cfss.uchicago.edu/index.html), Computing for the Social Sciences, a course taught by [Ben Soltoff](http://www.bensoltoff.com/) at the University of Chicago.

TO begin, we must load dependencies and import data to construct the workspace.

```{r message=FALSE}
library(tidyverse) # utility & visualization
library(knitr) # tables
library(kableExtra) # awesome tables

# Import data
load(file = "data/gun_deaths.rda")
```

## Understand the Variables
___
If the data were untidy, now would be the time to clean them. That is, we would want to ensure that each cell has exactly one value, each column one variable, and each row one observation. Thanfully, the instructor has cleaned the data ahead of time.

```{r}
glimpse(gun_deaths)
```

Aside from one `double` and `factor`, the data are comprised of `integer`s and `character`s.

What are the levels of the factor?

```{r}
levels(gun_deaths$education)
```

Alright, so the `education` variable is at the ordinal level.

## Number of Gun Deaths per Month
___
How many gun deaths occurred each month in the United States over the course of two years?

Before we tabulate and visualize the data, let's transform `month` into a factor with named levels.

```{r}
# Vector of month names
abbr_months <- c("Jan", "Fed", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")

# Transform month into a named factor
gun_deaths2 <- within(gun_deaths, {
  month <- factor(month, labels = abbr_months) 
})
```

Tables exhibit quantitative information more precisely than graphics. Algorithms prefer tabular data.

```{r}
# Count the rows, which represent deaths
count(gun_deaths2, month) %>%
  # Create the HTML table
  kable(caption = "Total Gun Deaths in the United States (2012-2014) per Month",
        col.names = c("Month", "Total Deaths")) %>%
  # Give the table some CSS love
  kable_styling(bootstrap_options = c("striped", "hover"))
```

From the table[^1], we see that the number of deaths appears constant. Aside from February, the total number of deaths for each month is above 8,000. It is important to remember that the table aggregates deaths for both years by month.

As humans, we tend to grasp patterns in visualizations better. The only number that jumps out to us after a quick glance at the above table is the leading digital, eight. Our minds blur the remaining digits as nonsensical. This is why $5.99 feels like five dollars and not six.

What does a bar chart tell us?

```{r}

```


## Dependencies
___
This activity was completed in RStudio.

```{r}
devtools::session_info()
```

[^1]: For more information on how to create awesome tables, check out Hao Zhu's [documentation](https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html).